### 2014-09-06 eval.tex
changes

@@

Our experiments and measurements 
were all conducted on an x86 64 Thinkpad W530 worksta- 
tion with eight 2.30GHz Intel Core i7-3610QM processors, 
16GB of RAM and 6M caches. The workstation runs version 12.04 of the Ubuntu Linux distribution, and has the Sun 
64-Bit 1.6.0\_26 Java virtual machine (JVM) installed.\footnote{We note that Sun JDK 1.7 does not support the transformation of dacapo applications with {\sf tamiflex}.}

@@

The $Races$ section shows the number of races detected by {\sf Recipe-s}, i.e., {\sf Recipe} without exploring un-executed paths; {\sf Recipe}, i.e., the full-fledged version; and {\sf RV}. Comparison between {\sf Recipe-s} and {\sf Recipe} reveals that {\sf Recipe} finds $>$100 more races, which demonstrates the strength of exploring  un-executed paths. Intuitively,  {\sf Recipe-s} predicts based on a single trace, while {\sf Recipe} predicts based on multiple traces containing different execution paths. We also compare between {\sf Recipe} and {\sf RV} version, as shown in the $Difference$ section, where $Diff$ shows the races found by {\sf Recipe} but missed by {\sf RV} while $Diff'$ shows the races found by {\sf RV} but missed by {\sf Recipe}.

@@

{\bf Summary of advantage and weakness\ } In general, our approach is flexible, which allows the relaxation of schedules and paths based on  fine-grained reasoning at the level of data values. As a result, it detects many more races compared to existing approaches. On the other hand, our approach is more heavyweight. An interesting topic for future research is how  to combine our approach with a lightweight approach with soundness guarantees like {\sf RV}. One option is to create a staged analysis, wherein {\sf RV} is first run. \tool\ is then left to run only on racy pairs not confirmed by {\sf RV}.
In this way, we can also tackle misses due to practical limitations of the constraint solver.

@@

\begin{figure}[htp]
\centering
\includegraphics[scale=0.7]{cases/Visio-bbuffer.pdf}
\caption{Relaxation of Inter-thread dependence}\label{fig:relax1}
\end{figure}

@@

\begin{figure}[htp]
\centering
\includegraphics[scale=0.7]{cases/Visio-msort.pdf}
\caption{Relaxation of Paths}\label{fig:relax2}
\end{figure}

>>>>

{\bf Prototype Implementation}  In our prototype implementation of \tool, we used the Z3 SMT solver for constraint resolution \cite{DeMoura:2008}. For static analysis, to discover unexecuted branches, we utilize the SOOT compiler framework \cite{Vallee-Rai:1999}.

For scalability, as well as to constrain memory footprint, we have applied several engineering optimizations. In specific, (i) events are stored in a database rather than in an in-memory buffer; and (ii) the analysis is split into several segments, between which we restart the VM and resume the last snapshot, such that the different logs are fused together at the end.

To address the scalability limitations of the constraint solver (Z3 in our case), we again apply a splitting optimization. The full trace is split into $N$ equal-length subtraces or ``time windows'' (in our experiments, $N=1,000$). The values of local and shared variables are then passed at the end of a given window as inputs to the next window. For lack of space, we omit the concrete details of this optimization, and instead refer the reader to Huang et al. \cite{HuangMR14}, who apply a similar optimization.

@@

Our experiments and measurements 
were all conducted on an x86 64 Thinkpad W530 worksta- 
tion with eight 2.30GHz Intel Core i7-3610QM processors, 
16GB of RAM and 6M caches. The workstation runs version 12.04 of the Ubuntu Linux distribution, and has the Sun 
64-Bit 1.6.0\_26 Java virtual machine (JVM) installed.
%\footnote{We note that Sun JDK 1.7 does not support the transformation of dacapo applications with {\sf tamiflex}.}

@@

The $Races$ section shows the number of races detected by {\sf Recipe-s}, i.e., {\sf Recipe} without exploring un-executed paths; {\sf Recipe}, i.e., the full-fledged version; and {\sf RV}. Comparison between {\sf Recipe-s} and {\sf Recipe} reveals that {\sf Recipe} finds $>$100 more races, which demonstrates the strength of exploring  un-executed paths. Intuitively,  {\sf Recipe-s} predicts based on a single trace, while {\sf Recipe} predicts based on multiple traces containing different execution paths. We also compare between {\sf Recipe} and {\sf RV} version, as shown in the $Difference$ section, where $Diff$ shows the races found by {\sf Recipe} but missed by {\sf RV} while $Diff'$ shows the races found by {\sf RV} but not {\sf Recipe}.

@@

To summarize, while our approach is able to relax schedules and paths based on  fine-grained reasoning at the level of data values, it also imposes higher overhead. An interesting topic for future research is how  to combine our approach with a lightweight approach with soundness guarantees like {\sf RV}. One option is staged analysis, wherein {\sf RV} is run. \tool\ is then left to run only on racy pairs not confirmed by {\sf RV}.
In this way, we can also tackle misses due to practical limitations of the constraint solver.

@@

\begin{figure}[htp]
\centering
\includegraphics[scale=0.45]{cases/Visio-bbuffer.pdf}
\caption{Relaxation of Inter-thread dependence}\label{fig:relax1}
\end{figure}

@@

\begin{figure}[htp]
\centering
\includegraphics[scale=0.45]{cases/Visio-msort.pdf}
\caption{Relaxation of Paths}\label{fig:relax2}
\end{figure}

### 2014-09-06 introduction.tex
intro
changes

However, we observe that such restrictions are unnecessary. For example, we can allow line {\tt 4} to receive its
value from line {\tt 1} as it has the same effect on thread local execution (i.e. the same branch is taken). 
However, the relaxation would allow us to execute lines {\tt 4} and {\tt 5} before line {\tt 2}, exposing 
the first missing race. We can also allow the trace to take the else branch such that the second race is also exposed. 
Note that even though the else branch was not executed originally, its effect can be precisely modeled as the value 
of $x$ is known from the original trace. As such, we can still perform sound analysis.

>>>>

However, we observe that such restrictions are unnecessary. For example, we can allow line {\tt 4} to receive its
value from line {\tt 1} as it has the same effect on thread local execution (i.e. the same branch is taken). 
This relaxation would allow us to execute lines {\tt 4} and {\tt 5} before line {\tt 2}, thereby exposing 
the first missing race. We can also allow the trace to take the else branch such that the second race is also exposed. 
Note that even though that branch was not executed originally, its effect can be precisely modeled as the value 
of $x$ is known from the original trace. As such, we can still perform sound analysis.

### 2014-09-06 overview.tex
changes

\subsection{Preliminaries}
To facilitate our presentation, we first introduce some terminologies used throughout this paper. Intuitively, a trace is a sequence of events recorded during the observation run.

{\bf Event\ } An event, $e=<t, id, inst>$, is a concrete representation of the runtime execution of a static instruction $inst$.

\begin{itemize}
\item $t$ refers to the thread that issues the event $e$, denoted as $t^e$.
\item  $id$ refers to the id associated with each event. The key property of  $id$ is {\em uniqueness}, i.e., any two events in the trace own different ids.  Unless otherwise specified, we use  the index of an event in the trace as its id, which satisfies the above property. Throughout this paper,  we denote an event as $e_{id}$ with the id as the subscript. %We may use the terms $e_{id}$ and $id$ interchangeably given their one-to-one correspondence, e.g., $t^{e_3}$ and $t^3$. 
\item $inst$ is the static instruction. The instructions are three-address instructions involving at most three operands, which modern compilers commonly support.  Specifically, we are interested in the types of instructions listed in Table~\ref{Ta:syntax}. When the variable does not appear on the left hand of an equation, such as $y$ in $x.f=y$, it may refer to a variable, a constant or event object creation expression $new (...)$.  The $bop$ stands for the binary operator, which may refer to $+, -, *, /, \%, \wedge, \vee$ in the assignment, or refer to $<, >, =, \wedge, \vee$ in the branch. The target of the branch event is not important in our scope, therefore, we may abbreviate the branch instruction as the boolean expression afterwards. The listed instructions suffice to represent all trace events of interest. This is because a concrete finite execution trace can be reduced to a straight-line loop-free call-free path program  (argument passing of the method call is modeled as assignments and the virtual call resolution is modeled as branches). This standard form of simplification preserves all the data-race-related information contained in the original trace. 
\end{itemize}

@@

{\bf Trace \ } A trace $\tau$ is a sequence of events. It can also be represented more comprehensively as,  $\tau=<\Gamma , \{\tau_{t_1}, \tau_{t_2}, \dots \tau_{t_n} \}, O, \theta>$.
\begin{itemize}
\item  $\Gamma$ refers to the set of threads involved in the trace, which include $t_1$, $t_2$, \dots, $t_n$.
\item   $\{\tau_{t_1}, \tau_{t_2}, \dots \tau_{t_n} \}$ denotes the event sequences produced by each thread.  
\item  $O$ is a function that assigns  to every event an integer value that denotes its scheduling order. For example, $e_i$ is scheduled earlier than $e_j$ if $O(e_i)<O(e_j)$.
\item  $\theta$ is a function that maps every variable to a value. For example, given an event $e$ executing the instruction $x=y+z$, the mapping recorded in the trace may be $\theta(x)=3$,  $\theta(y)=2$,  $\theta(z)=1$.  If the variable $x$ is defined in two events, the mapping for $x$ becomes ambiguous. To avoid the ambiguity, we conduct a pre-processing  (Section~\ref{sec:relax1}) that produces the SSA form of the trace, in which each variable is defined exactly once. 
\end{itemize}

@@

{\bf Race Condition\ } The potential race pair involves two accesses of the same shared location from different threads, where at least one access is a write. Given any potential race pair, e.g., ($e_2$, $e_5$), it is a real race if and only if the two accesses can occur at the same time (race condition), which is captured by the race condition constraint
$$
	O'(e_2) = O'(e_5)
$$.
The race condition constraint--- together with the other constraints --- guarantees the feasibility of the predicted race if a solution is found for the overall constraint system.

@@

\begin{itemize}
\item {\bf Control Flow Constraints\ } Each thread takes the same control flows (or branch decisions) to reproduce the events.
 For the running example (ignoring the statements in red), this yields:
$$
	\theta'(R_{\tt y}^4) > 2 \equiv {\bf true}
$$
That is, in the predicted trace $\tau'$, the value of variable {\tt y} read at $e_4$ should be greater than $2$.

\item {\bf Intra-thread Order Constraints\ } The events should follow the same thread-local order as in the original trace. This constraint is imposed by the fact that the two runs share the same instruction sequence and take the same control flows. 
We yield the following formula:
$$
	O'(e_1) < O'(e_2) < O'(e_3) \wedge O'(e_4) < O'(e_5)
$$
That is, in the predicted trace $\tau'$, the three events from $T_1$ are ordered in the same way as in $\tau$, and similarly, the branch event $e_4$ in $T_2$ still executes before the event $e_5$ inside the branch body.

@@

{\bf Constraint Solving\ } The formuals from the different encoding steps are conjoined and sent to the off-the-shelf solver such as {\sf z3}.  The race under analysis is real if the solver returns a solution, which includes the scheduling order and the values for the variables.

{\bf Highlights \ } In this example, our analysis finds two more pairs of races than existing analyses~\cite{yannis, jeff}, which include $(e_2, e_5)$ and $(e_2, e_7)$.
The key insight for finding the first race is, we do not require $e_4$ to read from $e_3$ to retain the original value, as what existing approaches do, rather, we allow $e_4$ to read from $e_1$ while still preserving the feasibility of the branch. This relaxation allows the events after $e_4$ to run concurrently with the events before $e_3$, leading to the detection of the first race.  The key insight for finding the second race is, we explore the unexplored branches and ask the solver to compute a feasible schedule that exercises them. The neighboring branches are likely to contain racy events if the executed branches contain racy events.

>>>>

\subsection{Preliminaries}
We first introduce some notation used throughout this paper. Intuitively, a trace is a sequence of events recorded during the observed run. An event, $e=<t, id, inst>$, is a concrete representation of the runtime execution of a static instruction $inst$, such that (i) $t$ refers to the thread that issues the event $e$ (denoted $t^e$); (ii) $id$ is a unique identifier of the event (unless spcified otherwise, we use the index of an event in the trace as its id); and (iii) $inst$ is a static three-address instruction (involving at most three operands, which modern compilers commonly support).

For this paper, we utilize the instructions listed in Table~\ref{Ta:syntax}. When the variable does not appear on the left hand of an equation, such as $y$ in $x.f=y$, it may refer to a variable, a constant or event object creation expression $new (...)$.  The $bop$ stands for the binary operator, which may refer to $+, -, *, /, \%, \wedge, \vee$ in the assignment, or refer to $<, >, =, \wedge, \vee$ in the branch. The target of the branch event is not important in our scope, therefore, we may abbreviate the branch instruction as the boolean expression afterwards. The listed instructions suffice to represent all trace events of interest. This is because a concrete finite execution trace can be reduced to a straight-line loop-free call-free path program  (argument passing of the method call is modeled as assignments and the virtual call resolution is modeled as branches). This standard form of simplification preserves all the data-race-related information contained in the original trace.

@@

{\bf Trace Representation \ } In the following, we assume that trace $\tau$ is represented as $\tau=<\Gamma , \{\tau_{t_1}, \tau_{t_2}, \dots \tau_{t_n} \}, O, \theta>$, where
\begin{itemize}
\item  $\Gamma$ refers to the set of threads involved in the trace, which include $t_1$, $t_2$, \dots, $t_n$;
\item   $\{\tau_{t_1}, \tau_{t_2}, \dots \tau_{t_n} \}$ denotes the event sequences produced by each thread;
\item  $O$ is a function that assigns  to every event an integer value that denotes its scheduling order: $e_i$ is scheduled earlier than $e_j$ iff $O(e_i)<O(e_j)$; and
\item  $\theta$ is a function that maps every live variable to a value. 
%For example, given an event $e$ executing the instruction $x=y+z$, the mapping recorded in the trace may be $\theta(x)=3$,  $\theta(y)=2$,  $\theta(z)=1$.  
\end{itemize}
To ensure the integrity of $\theta$ as well simplify downstream encoding steps performed by the analysis already at the syntactic level, we rewrite the raw input trace into SSA form, such that each variable is defined exactly once.

@@

{\bf Race Condition\ } The potential race pair involves two accesses of the same shared location from different threads, where at least one access is a write. Given any potential race pair, e.g., ($e_2$, $e_5$), it is a real race if and only if the two accesses can occur at the same time (race condition), which is captured by the race condition constraint: 
$O'(e_2) = O'(e_5)$.
The race condition constraint--- together with the other constraints --- guarantees the feasibility of the predicted race if a solution is found for the overall constraint system.

@@

\begin{itemize}
\item {\bf Control Flow Constraints\ } Each thread takes the same control flows (or branch decisions) to reproduce the events.
 For the running example (ignoring the statements in red), this yields: $\theta'(R_{\tt y}^4) > 2 \equiv {\bf true}$.
That is, in the predicted trace $\tau'$, the value of variable {\tt y} read at $e_4$ should be greater than $2$.

\item {\bf Intra-thread Order Constraints\ } The events should follow the same thread-local order as in the original trace. This constraint is imposed by the fact that the two runs share the same instruction sequence and take the same control flows. 
We obtain the  formula: $O'(e_1) < O'(e_2) < O'(e_3) \wedge O'(e_4) < O'(e_5)$.
That is, in the predicted trace $\tau'$, the three events from $T_1$ are ordered in the same way as in $\tau$, and similarly, the branch event $e_4$ in $T_2$ still executes before the event $e_5$ inside the branch body.

@@

{\bf Constraint Solving\ } The formuals from the different encoding steps are conjoined and sent to an off-the-shelf solver such as {\sf z3}.  The race under analysis is guaranteed to be real if the solver returns a solution, which includes the scheduling order and the values for the variables.

{\bf Highlights \ } In this example, our analysis finds two more pairs of races beyond existing analyses~\cite{yannis, jeff}: $(e_2, e_5)$ and $(e_2, e_7)$.
For the first race, we do not require $e_4$ to read from $e_3$ to retain the original value, as in approaches. Rather, we allow $e_4$ to read from $e_1$ while still preserving the feasibility of the branch. This relaxation allows the events after $e_4$ to run concurrently with the events before $e_3$, leading to the detection of the first race.  For the second race, we cover unexecuted branches and ask the solver to compute a feasible schedule that exercises them. The neighboring branches are likely to contain racy events if the executed branches contain racy events.

### 2014-09-06 related.tex
changes

Initial attempts to address the challenge of race detection focused on built-in synchronization primitives \cite{eraserPaper,Agarwal,vonPraun:2001,Choi:2002}. These include locks as well as the wait/notify and start/join scheduling controls. Notable among these efforts is \emph{lockset} analysis, which considers only locks \cite{eraserPaper}. Because the derived constraints are partial, permitting certain infeasible event reorderings, lockset analysis cannot guarantee soundness \cite{Naik:2006}.

A different tradeoff is struck by the \emph{happens-before (HB)} approach \cite{Christiaens:2001,Dinning:1990,Mellor-Crummey:1991}, which is inspired by and based on Lamport's HB relation \cite{Lamport}. In this style of analysis, all synchronization primitives are accounted for, though reordering constraints are conservative . As an example, HB inhibits reordering of two synchronized blocks governed by the same lock.

@@

Recently there have been successful attempts to relax HB constraints. Among these are hybrid analysis \cite{XXX}, which permits both orderings of lock-synchronized blocks, as well as the \emph{Universal Causal Graph (UCG)} representation~\cite{ucg}, which also enables both orderings but only if these are consistent with wait/notify- and start/join-induced constraints.

@@

The first to propose PTA are Smagardakis et al. \cite{Smaragdakis:2012}. In their original work, both synchronization constraints and inter-thread dependencies are preserved, where inter-thread dependencies are respected by 
only allowing reorderings that leave the dependence structure exhibited by the original trace in tact. This ensures that the values of shared memory locations remain the same, which secures the soundness argument, though maximality is not guaranteed.

>>>>

Initial attempts to address the challenge of race detection focused on built-in synchronization primitives \cite{eraser,SasturkarAWS05,vonPraun:2001,Choi:2002}. These include locks as well as the wait/notify and start/join scheduling controls. Notable among these efforts is \emph{lockset} analysis, which considers only locks \cite{eraser}. Because the derived constraints are partial, permitting certain infeasible event reorderings, lockset analysis cannot guarantee soundness \cite{Naik:2006}.

A different tradeoff is struck by the \emph{happens-before (HB)} approach \cite{Christiaens,Dinning:1990,Mellor-Crummey:1991}, which is founded on Lamport's HB relation \cite{Lamport}. In this style of analysis, all synchronization primitives are accounted for, though reordering constraints are conservative . As an example, HB inhibits reordering of two synchronized blocks governed by the same lock.

@@

Recently there have been successful attempts to relax HB constraints. Among these are hybrid analysis \cite{hybrid}, which permits both orderings of lock-synchronized blocks, as well as the \emph{Universal Causal Graph (UCG)} representation~\cite{ucg}, which also enables both orderings but only if these are consistent with wait/notify- and start/join-induced constraints.

@@

The first to propose PTA are Smagardakis et al. \cite{yannis}. In their original work, both synchronization constraints and inter-thread dependencies are preserved, where inter-thread dependencies are respected by 
only allowing reorderings that leave the dependence structure exhibited by the original trace in tact. This ensures that the values of shared memory locations remain the same, which secures the soundness argument, though maximality is not guaranteed.
